## Wrangling verbs

The vast majority of data cleaning tasks can be handled with five verbs:

- `filter()` Pick observations by their values.
- `arrange()` Reorder the rows.
- `select()` Pick variables by their names.
- `mutate()` Create new variables with functions of existing variables.
- `summarise()` Collapse many values down to a single summary.

Oh, and if you use the pipe, `%>%`, you can string several of these actions together to make a data pipeline.

You can find a data wrangling cheatsheet [here](https://github.com/rstudio/cheatsheets/raw/master/source/pdfs/data-transformation-cheatsheet.pdf).


### Filter

Filter is used to pick some cases or observations (rows) in the data. Here's an example for the `pokemon` data. We will choose just the Pokemons that are "grass" type (use the `type1` variable). 

```{r, eval = FALSE}
## remotes::install_github("schochastics/Rokemon")
library(Rokemon)

readr::write_csv(
  x = Rokemon::pokemon %>% 
    dplyr::select(pokedex_number, name, 
                  type1, type2, 
                  # generation, is_legendary,
                  height_m, weight_kg,
                  attack, defense
                  # hp, speed,
                  # sp_attack, sp_defense
                  ),
  file = "./data/pokemon_demo.csv")

readr::write_csv(
  x = Rokemon::pokemon %>% dplyr::select(-abilities),
  file = "./data/pokemon_full.csv")
```

```{r echo=TRUE}
pokemon %>% filter(type1 == "grass")
```

*Try it!* Do these exercises:

- Filter the `pokemon` data to pick only Pokemons that are "fire" type (use the `type2` variable)
- Filter the Pokemons that are more than 150 units in `attack` (HINT: you can use `>` or `<` if the variable is just made up of numbers)
- Filter the Pokemons that are either "fire" in `type1` **or** "water" in `type1` (HINT: You need to use the operator `%in%` like this `%in% c("fire","water"))`
- Filter the POkemons that are "fire" in `type1` **and** "water" in `type2` (HINT: You can add additional conditions in the `filter` function by adding an extra comma like `filter(type1 == "fire", ...)`)

```{r, eval = FALSE}
pokemon %>% filter(type2 == "fire")
pokemon %>% filter(attack > 150)
pokemon %>% filter(type1 %in% c("fire", "water"))
pokemon %>% filter(type1 == "fire", type2 == "water")
```

```{r filter, exercise=TRUE}

```

### Select

Select is used to pick some variables in the data. Here's an example for the `pokemon` data. We will select just the variables `name` and `attack`.

```{r echo=TRUE}
pokemon %>% select(name, attack) 
```

*Try it!* Select the variables `name`, `type1`, `attack` and `defense`.

```{r select, exercise=TRUE}

```

### Arrange

`arrange` sorts the data by values in one of the columns. Here's an example which also involves selecting a subset of variables.

```{r echo=TRUE}
pokemon %>% 
  select(name, attack) %>%
  arrange(desc(attack))
```

Note that `desc` arranges in descending order. 

*Try it!* Arrange Pokemons by `attack`, in increasing order.

```{r arrange, exercise=TRUE}

```

Arrange is mostly used to get quick views of the numbers.

### Mutate

I love the name mutate! It means to create new variables, or modify existing ones. For the `pokemon` dataset, we might be interested in examining the sum of both `attack` and `defense` as a simple way of summairsing a Pokemon's ability. Here's how we can create this variable:

```{r echo=TRUE}
pokemon %>% 
  mutate(total = attack + defense)
```

*Try it!* Compute a new variable called `bmi` (body mass index), which is computed as:

$$bmi = \frac{\text{weight}}{\text{height}^2}.$$

```{r mutate, exercise=TRUE}

```

### Summarise

Summarise is the workhorse function. It takes columns of the data, and reduces them to single numbers. It is most useful when we want to compute statistics for our. 

For the `pokemon` data, we might be interested in computing the mean, minimum and maximum of `attack` value of all Pokemons. 

```{r echo=TRUE}
pokemon %>% 
  summarise(
    min = min(attack),
    mean = mean(attack),
    max = max(attack)
  )
```
S
However, the `attack` value might be dependent on the type of the Pokemons, so we can compute this separately for each of the `type1` categories. Which type of Pokemon has the highest average `attack`?

```{r echo=TRUE}
pokemon %>% 
  group_by(type1) %>% 
  summarise(mean = mean(attack))
```

If it helps, we can use `arrange` to better help us to answer this question. 

```{r}
pokemon %>% 
  group_by(type1) %>% 
  summarise(mean = mean(attack)) %>% 
  arrange(desc(mean))
```

*Try it!* Compute the mean and standard deviation for (a) `attack`, (b) `defense`.

```{r summarise, exercise=TRUE}

```
